\texttt{\texttt{NeuralNetworkVisualization}}

### 2019-09-22 paper_neural_network.Rmd
Removed small typos.

An approach from classic statistics is the interpretation of marginal effects. Neural networks are black box models, as the internal observation of their functioning is not possible. In order to determine the effect of predictors, one does not aim at opening the black box, but to observe the behavior of the black box at the change of the input. One of the methods to achieve this goal is the partial dependence plot [@Friedman.2001]. In order to implement this approach, we created the package \texttt{NeuralNetworkVisualization} in the software R. This paper aims to offer a dynamic approach for visualizing the partial dependence plots with R Shiny. This report is structured as follows: The first section assesses the theory which lays the foundation for the implementation of the code. The second chapter introduces and examines the created package. The next section explains the code structure. For illustration, the fourth chapter then gives examples. The last section concludes the realization of partial dependence plots.

@@

# Introducing the \texttt{NeuralNetworkVisualization}-package
We created a package for R which produces partial dependence plots in different presentations and for different types of data. The technical details of this package will be discussed in the next section.

@@

## \texttt{NeuralNetwork}
Implementing a package that is able to create partial dependence plots requires having the ability to fit neural networks easily. \texttt{neuralnet} is one of the most common R packages used for fitting neural networks but it has some minor inconveniences. First of all, factor predictor variables have to be expanded to a set of dummy variables. In the \texttt{neuralnet} package this has to be done manually before fitting the model using the \texttt{class.ind} function from the \texttt{nnet} package. Furthermore, if the range of the predictors is different, scaling or data normalization has to be performed. Again, in the \texttt{neuralnet} function this has to be done before fitting the model. Additionally, we need to refit the same model multiple times for creating the bootstrap confidence intervals for the partial dependence plots. Thus, our package functionality needs to remember all specifications for the network provided by the user. Hence, we decided to write a wrapper class around the \texttt{neuralnet} function.

@@

## \texttt{prepare\_data}
Data preparation for plotting is hidden from the end user. The \texttt{prepare\_data} function is responsible for the plotting data creation as explained by the algorithm in chapter 2.3. Besides preparing the plotting data, the function also adds the lower and upper levels for the confidence intervals if desired by the user. If no data was stored in the fitted model via the \texttt{options} parameter, every visualization will use this particular to obtain the necessary data for plotting. Moreover, if the \texttt{scale} parameter of the \texttt{NeuralNetwork} has been set, the data will be descaled back to the original data range for plotting. The functionality for the data preparation is stored in the **plotting_data_preparation.R** file.

If the user chooses to add cofidence intervals, for each bootstrap iteration the neural network will be refitted with sampled data. This procedure might take some time, since fitting a neural network depends on the available data, the number of hidden layers and the threshold for the minimization of the prediction error. As mentioned, we added the possibility to use parallel computing to decrease running time of the bootstrap procedure. Additionally, it is possible to store the plotting data directly in the \texttt{NeuralNetwork} class via the options parameter. This reduces waiting time for the desired visualizations.

@@

First, let us fit the neural network using the \texttt{NeuralNetwork} class. Compared to the \texttt{neuralnet} package less data preprocessing has to be done by the user. He or she does not have to consider factor columns that should be expanded into dummy variables, and scaling does not need to be done manually. By providing the \texttt{options} parameter, the data is used for visualizing the partial dependencies is directly computed. The model fitting takes time, since 4000 neural networks are refitted by the bootstrap procedure. The resulting data is stored inside the \texttt{NeuralNetwork}.

@@

Let us consider the partial dependence plots with bootstrap confidence intervals to capture the uncertainty of the effect. The thick line represents the average partial dependence and the lower as well as the upper bound of the respective quantiles. We can just use the stored data to create the partial dependence plot with the \texttt{probs} and \texttt{nrepetitions} parameters specified as in the \texttt{options} parameter where the model has been fitted.

@@

Now that the confidence intervals have been added to the partial dependence plots, the interpretation can additionally capture the uncertainty of the effect of a change of the predictor on the response category. Furthermore, if we want to look at only one predictor, we can just specify the \texttt{predictors} parameter. Instead of using the data already stored, we  compute the data for plotting directly.

@@

# Conclusion
This paper gave a brief introduction into the concept of interpretability for neural network model featuring  partial dependence plots. These plots have been examined and implemented in our \texttt{NeuralNetworkVisualization} package.

@@

\newpage
#Statement of Ownership
Verification of examination registration in FlexNow  
Name: Ms Jacqueline Dorothea Seufert (Matriculation No.: 21800321)  
Name: Mr Alex Afanasev (Matriculation No.: 21800878)  
Semester: SoSe19  
Degree Course: Angewandte Statistik (Master of Science)  
Module: Advanced Statistical Programming with R [M.WIWI-QMW.0011]  
Exam: M.WIWI-QMW.0011.Mp: Advanced Statistical Programming with R  
(Präsentation (ca. 20 Minuten) mit schriftlicher Ausarbeitung (6-8 Seiten))  
Exam Date: 23.09.2019  
Lecturer: Prof. Dr. Thomas Kneib  
*Declaration*  
We hereby declare that we have produced this work independently and without outside
assistance, and have used only the sources and tools stated.
We have clearly identified the sources of any sections from other works that we have quoted
or given in essence.
We have complied with the guidelines on good academic practice at the University of
Göttingen.
If a digital version has been submitted, it is identical to the written one.
we are aware that failure to comply with these principles will result in the examination
being graded “nicht bestanden”, i.e. failed.  
\bigskip
\bigskip
Göttingen, 18th September 2019\
\bigskip
\bigskip
Alex Afanasev\
\bigskip
\bigskip
Jacqueline Dorothea Seufert\
\newpage

>>>>

An approach from classic statistics is the interpretation of marginal effects. Neural networks are black box models, as the internal observation of their functioning is not possible. In order to determine the effect of predictors, one does not aim at opening the black box, but to observe the behavior of the black box at the change of the input. One of the methods to achieve this goal is the partial dependence plot [@Friedman.2001]. In order to implement this approach, we created the package \texttt{NeuralNetworkVisualization} in the software R. This package aims to offer a dynamic approach for visualizing the partial dependence plots with R Shiny. This report is structured as follows: The second section assesses the theory which lays the foundation for the implementation of the code. The third chapter introduces and examines the created package. The next section explains the code structure. For illustration, the fifth chapter then gives examples. The last section concludes the realization of partial dependence plots.

@@

# Introducing the \texttt{NeuralNetworkVisualization} - package
We created a package for R which produces partial dependence plots in different presentations and for different types of data. The technical details of this package will be discussed in the next section.

@@

## \texttt{NeuralNetwork}
Implementing a package that is able to create partial dependence plots requires having the ability to fit neural networks easily. \texttt{neuralnet} is one of the most common R packages used for fitting neural networks but it has some minor inconveniences. First of all, factor predictor variables have to be expanded to a set of dummy variables. This has to be done manually before fitting the model and can be accomplished using the \texttt{class.ind} function from the \texttt{nnet} package. Furthermore, if the range of the predictors is different, scaling or data normalization has to be performed. Again, in the \texttt{neuralnet} function this has to be done before fitting the model. Additionally, we need to refit the same model multiple times for creating the bootstrap confidence intervals for the partial dependence plots. Thus, our package functionality needs to remember all specifications for the network provided by the user. Hence, we decided to write a wrapper class around the \texttt{neuralnet} function.

@@

## \texttt{prepare\_data}
Data preparation for plotting is hidden from the end user. The \texttt{prepare\_data} function is responsible for the plotting data creation as explained by the algorithm in chapter 2.3. Besides preparing the plotting data, the function also adds the lower and upper levels for the confidence intervals if desired by the user. If no data was stored in the fitted model via the \texttt{options} parameter, every visualization will use this function to obtain the necessary data for plotting. Moreover, if the \texttt{scale} parameter of the \texttt{NeuralNetwork} has been set, the data will be descaled back to the original data range for plotting. The functionality for the data preparation is stored in the **plotting_data_preparation.R** file.

If the user chooses to add cofidence intervals, the neural network will be refitted with sampled data for each bootstrap iteration. This procedure might take some time, since fitting a neural network depends on the available data, the number of hidden layers and the threshold for the minimization of the prediction error. As mentioned, we added the possibility to use parallel computing to decrease running time of the bootstrap procedure. Additionally, it is possible to store the plotting data directly in the \texttt{NeuralNetwork} class via the \texttt{options} parameter. This reduces waiting time for the desired visualizations.

@@

First, let us fit the neural network using the \texttt{NeuralNetwork} class. Compared to the \texttt{neuralnet} package less data preprocessing has to be done by the user. He or she does not have to consider factor columns that should be expanded into dummy variables, and scaling does not need to be done manually. By providing the \texttt{options} parameter, the data that is used for visualizing the partial dependencies is directly computed. The model fitting takes time, since 4000 neural networks are refitted by the bootstrap procedure. The resulting data is stored inside the \texttt{NeuralNetwork}.

@@

Let us consider the partial dependence plots with bootstrap confidence intervals to capture the uncertainty of the effect. The thick line with points represents the average partial dependence and the thiner lines around it the lower as well as the upper bound from the confidence interval. We can use the stored data to create the partial dependence plot with the \texttt{probs} and \texttt{nrepetitions} parameters specified as in the \texttt{options} parameter where the model has been fitted.

@@

Now that the confidence intervals have been added to the partial dependence plots, the interpretation can additionally capture the uncertainty of the effect of a change of the predictor on the response category. Furthermore, if we want to look at only one predictor, we can just specify the \texttt{predictors} parameter. Instead of using the already stored data, we compute the data for plotting directly.

@@

# Conclusion
This paper gave a brief introduction into the concept of interpretability for neural network models featuring  partial dependence plots. These plots have been examined and implemented in our \texttt{NeuralNetworkVisualization} package.

@@

\newpage
# Statement of Ownership

Verification of examination registration in FlexNow  
Name: Ms Jacqueline Dorothea Seufert (Matriculation No.: 21800321)  
Name: Mr Alex Afanasev (Matriculation No.: 21800878)  
Semester: SoSe19  
Degree Course: Angewandte Statistik (Master of Science)  
Module: Advanced Statistical Programming with R [M.WIWI-QMW.0011]  
Exam: M.WIWI-QMW.0011.Mp: Advanced Statistical Programming with R  
(Präsentation (ca. 20 Minuten) mit schriftlicher Ausarbeitung (6-8 Seiten))  
Exam Date: 23.09.2019  
Lecturer: Prof. Dr. Thomas Kneib

*Declaration*

We hereby declare that we have produced this work independently and without outside assistance, and have used only the sources and tools stated. We have clearly identified the sources of any sections from other works that we have quoted or given in essence. We have complied with the guidelines on good academic practice at the University of Göttingen. If a digital version has been submitted, it is identical to the written one. we are aware that failure to comply with these principles will result in the examination being graded “nicht bestanden”, i.e. failed.

\bigskip
\bigskip
Göttingen, 18th September 2019\

\bigskip
\bigskip
Alex Afanasev\

\bigskip
\bigskip
Jacqueline Dorothea Seufert\
\newpage
